Physics-based protein structure prediction and design using the confinement method

### 2013-01-09 Paper/paper.tex
Jan10 3 Draft

The calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular
conformations as well as a computationally expensive task with existing
methods. In this work, the confinement approach is used to calculate absolute
free energies of biomolecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. The free energy calculated can be decomposed into a per
residue contribution in an 
approximate way. Per residue free energy allows us
to identify the reason behind conformational preferences in biomolecules. Through out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preferences and per residue free energy decomposition set the stage for the use of this 
method in protein design.

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. The ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be
the molecular cause of a range of human genetic disorders. For example a misfolded protein known as
a
prion can misfold correctly folded proteins when entering a healthy organism.
\Alberto{I'm not sure I understand this here} \Arijit{Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein.} These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change becomes difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescales involved in
such conformational transitions may
be beyond the sampling ability of classical molecular dynamics simulation.  Generally, in order to
calculate the free energy a pathway or reaction coordinate is needed with several intermediate structures.
Methods like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and require a large computational effort. The idea of a reaction coordinate became
even more
complicated in the case of protein folding.  As the free energy landscape for protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}, the molecule 
can get trapped in an energy well during conformational sampling. Even with 
advanced methods like replica exchange molecular dynamics, the three dimensional structure may be
trapped in a local minima for a considerable time, giving rise to nonconverged distributions where
misfolded states might be selected as native like
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

The calculation of free energy has been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

@@

\begin{figure}
\includegraphics[width=7 in,height=5.8 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Orientation of residues of different sidechains in $4 \beta + \alpha$ and $3 \alpha$ form
A. Orientation of Hydrophobic residues Leu-5 and Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Hydrophobic Ala-26
D. Hydrophobic Val-54.
E. Hydrophilic Glu-14.
F. The plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak 
stabilize $4 \beta + \alpha$. 
G. Hydrophilic Glu-56 .
H. Orientation of Hydrophic Ile-49.
I. With GA95 sequence, Ile-49 have hydrophic interaction with both Leu-20 and Ile-30.
J. Larger Phe-30 in GB sequence, move towards the protein surface in $4 \beta + \alpha$.
K. Tyr-45 form H-bond with Asp-47 in $4 \beta + \alpha$.}  
\end{figure}

>>>>

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is particularly difficult and computationally expensive task with existing
methods. In this work, confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. The free energy calculated can be decomposed into per residue contribution in a 
approximate way. Per residue free energy allow us
to identify the reason behind conformational preference of the biomolecules. Through out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference and per residue free energy decomposition set the stage for the use of this 
method in protein design.

@@

In computational structural biology there are numerous cases where free energy between well defined
states are necessary. Examples ranges from small to large conformational change of protein due to
ligand binding, change of pH etc \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also play an important role in case of protein
folding. The ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting point or unfolding
it by changing different condition most proteins will finally assume the same structure. It is very
important for proteins to achieve their native conformation since defects in protein folding may be
the molecular cause of a range of human genetic disorder. For example a misfolded protein known as
prion when enters any healthy organism it converts properly folded protein into misfolded one.
Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein. These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change become difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescale involved in such tranformation may
be beyond timescale of direct classical molecular dynamics simulation.  Generally, in order to
calculate free energy a pathway or reaction coordinate is created with the intermediate structures.
Method like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and require large computational effort. The idea of reaction coordinate became more
complicated in case of protein folding.  As free energy landscape of the protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}
it can be trapped in a energy well during its visit in the conformational space. Even with the
advanced method like replica exchange molecular dynamics the three dimensional structure may be
trapped in a local minima for a considerable time and trapped in a local energy minima
and one can mistakenly identify the misfolded state as the native state.
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy has been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}, . In recent years methods like Refrence system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008}, Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} have came out for calculation of free energy method. 
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that free energy is a state function and 
thus independent of path. It uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraint to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy of conformation of pair of proteins with similar sequence but completely
different fold was calculated. Encouraged by the result, we applied the method of different targets
of CASP9 competition and tested whether it can predict the most stable conformation of the protein based 
on its free energy. The remarkable ability of confinement method to identify residues which 
stabilizes/destabilizes a particular conformation from per residue free energy analysis set the stage
for future application of this method in protein design.

@@

\begin{figure}
\includegraphics[width=6.5 in,height=5.4 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Orientation of residues of different sidechains in $4 \beta + \alpha$ and $3 \alpha$ form
A. Orientation of Hydrophobic residues Leu-5 and Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Hydrophobic Ala-26
D. Hydrophobic Val-54.
E. Hydrophilic Glu-14.
F. The plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak 
stabilize $4 \beta + \alpha$. 
G. Hydrophilic Glu-56 .
H. Orientation of Hydrophic Ile-49.
I. With GA95 sequence, Ile-49 have hydrophic interaction with both Leu-20 and Ile-30.
J. Larger Phe-30 in GB sequence, move towards the protein surface in $4 \beta + \alpha$.
K. Tyr-45 form H-bond with Asp-47 in $4 \beta + \alpha$.}  
\end{figure}
